Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-11 arxiv/diff_cov.tex
sec 3

	\noindent (i) The variance equation \eqref{eq_variance_2task} satisfies the following estimate with high probability:
			\begin{align}\label{lem_cov_shift_eq}
				\bigabs{p^{-1}\bigtr{\Sigma^{(2)} \bigbrace{(n_1 + n_2){\hat{\Sigma}^{-1}} - (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}}} \le  p^{-c_{\varphi}}, %\norm{\Sigma^{(2)}},
			\end{align}
			where $a_1$ and $a_2$ are the solutions of the following self-consistent equations
			\begin{align}
				a_1 + a_2 = 1- \frac{p}{n_1 + n_2},\quad a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} = \frac{n_1}{n_1 + n_2}. \label{eq_a12extra}
			\end{align}
	\noindent (ii) The bias equation \eqref{eq_bias_2task} satisfies the following limit with high probability for any fixed vector $w\in\real^p$ with unit norm,
			\begin{align}\label{lem_cov_derv_eq}
				\bigabs{w^{\top} \Sigma^{(1)} \bigbrace{(n_1+n_2)^2\hat{\Sigma}^{-1} \Sigma^{(2)} \hat{\Sigma}^{-1} - {\Sigma^{(2)}}^{-1/2} V {\frac{a_3 \Lambda^2 + (a_4 + 1)\id}{(a_1 \Lambda^2 + a_2\id)^2}} V^{\top} {\Sigma^{(2)}}^{-1/2}} \Sigma^{(1)} w} \le  p^{-c_{\varphi}},
			\end{align}
				where $a_{3}$ and $a_4$ are the solutions of the following self-consistent equations % with $b_k = \frac1{p}\sum_{i=1}^p \frac{\lambda_i^{2k}} {(\lambda_i^2 a_1 + a_2)^2}$, for $k = 0, 1, 2$:
			\begin{align}\label{eq_a34extra}
				a_3 + a_4 = \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2}, \ \ 
				a_3 + \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 (a_1 a_4 - a_2 a_3)}{(\lambda_i^2 a_1 + a_2)^2} = \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^{2}}.
%				\left(\frac{\rho_1}{a_1^{2}} -  b_2  \right)\cdot  a_3 -  b_1 \cdot  a_4 = b_1,\quad \left(\frac{\rho_2}{a_2^{2}}-  b_0\right)\cdot  a_4 - b_1 \cdot  a_3
%				= b_0.
			\end{align}
\end{theorem}
Our result extends Fact \ref{fact_tr} on the inverse of a single sample covariance matrix.
When $n_1$ is zero, we solve equation \eqref{eq_a12extra} and obtain $a_1 = 0$ and $a_2 = (n_2-p) / n_2$, and apply these solutions to equation \eqref{lem_cov_shift_eq}.

@@

\paragraph{Bias asymptotic limit.}
For the bias limit in equation \eqref{lem_cov_derv_eq}, we show that it is governed by the derivative of $(Y - z\id)^2$ with respect to $z$ at $z = 0$.
First, we reduce the empirical bias term in equation \eqref{lem_cov_derv_eq} to $Y$
\begin{align}\label{calculate G'}
	(n_1 + n_2)^2 \hat{\Sigma}^{-1}\Sigma^{(2)}\hat{\Sigma}^{-1} = {\Sigma^{(2)}}^{-1/2} V Y^{-2} V^{\top} {\Sigma^{(2)}}^{-1/2}.
\end{align}
Let $\cal G(z)$ denote the resolvent of $Y$.
Our key observation is that $\frac{\dd{\cal G(z)}}{\dd z} = - \cal G^2(z)$.
Hence, provided that limit of $(Y - z\id)^{-1}$ is $(a_1(z) \Lambda^2 + (a_2(z) - z) \id)^{-1}$ near $z = 0$, the limit of $\frac{\dd{\cal G(0)}}{\dd z}$ satisfies that
\begin{align}\label{cal G'0}
	\frac{\dd \cal G(0)}{\dd z} = \frac{-\frac{\dd a_1(0)}{\dd z}\Lambda^2 - (\frac{\dd a_2(0)}{\dd z} - 1)\id}{(a_{1}(0)\Lambda^2 + a_{2}(0)\id_p)^2}.
\end{align}
For the derivative of $a_1(z)$ and $a_2(z)$, we take the derivative on both sides of equation \eqref{selfomega_a}.
%\begin{align*}
%	\frac{\dd a_1(z)}{\dd z} + \frac{\dd a_2(z)}{\dd z} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2},
%	\frac{\dd a_1(z)}{\dd z} + \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{\lambda_i^2 (a_1'(z) a_2 - a_2'(z) a_1)}{(\lambda_i^2 a_1 + a_2)^2} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^2}
%\end{align*}
Let $a_3 = - \frac{\dd a_1(0)}{\dd z}$ and $a_4 = - \frac{\dd a_2(z)}{\dd z}$.
%then taking implicit differentiation of equation \eqref{selfomega_a}
One can verify that $a_3$ and $a_4$ satisfy the self-consistent equations \eqref{eq_a34extra} (details omitted).
Applying equation \eqref{cal G'0} to equation \eqref{calculate G'}, we obtain that the asymptotic limit of the bias term.
%\begin{align}
%& n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}) }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
%&\approx  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\frac{a_3\Lambda^2 +(1+ a_4)\id_p}{(a_{1}\Lambda^2 + a_{2}\id)^2}V^\top \Sigma_2^{-1/2}\Sigma_1^{1/2}w= w^{\top} \Pi_\bias w, \label{calculatePibias}
%\end{align}
%where in the last step we used $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $V \Lambda^2 V^\top=M^\top M$. This concludes equation \eqref{lem_cov_derv_eq}.

As a remark, for $\frac{\dd \cal G(z)}{\dd z}$ to stay close to its limit at $z = 0$, we not only need the limit of $\cal G(0)$, but also the limit of $\cal G(z)$ within a small neighborhood of $0$.
This is why we consider $Y$'s resolvent matrix as opposed to the Stieljes transform of its empirical spectral distribution.

\paragraph{Schur complement and self-consistent equations.}
We briefly describe how to derive the matrix limit $\Gi(z)$.
First, we consider the case where $Z^{(1)}$ and $Z^{(2)}$ are both multivariate Gaussian random matrices.
By rotational invariance, we have that $Z^{(1)} U$ and $Z^{(2)} V$ are both still multivariate Gaussian random matrices.
Next, we use the Schur complement formula to deal with the resolvent $G(z)$.
We show that $G(z)$'s diagonal entries satisfy a set of self-consistent equations in the limit, leading to equation \eqref{selfomega_a}.
$G(z)$'s off-diagonal entries are approximately zero using standard concentration bounds.
Finally, we extend our result to general random matrices under constant moment condition.
We prove an anisotropic local law using recent developments in random matrix theory \cite{erdos2017dynamical,Anisotropic}.
A complete proof of Theorem \ref{thm_main_RMT} can be found in Appendix \ref{appendix RMT}.

>>>>

	\noindent (i) The variance equation \eqref{eq_variance_2task} satisfies the following estimate with high probability:
			\begin{align}\label{lem_cov_shift_eq}
				\bigabs{p^{-1}\bigtr{\Sigma^{(2)} \bigbrace{(n_1 + n_2){\hat{\Sigma}^{-1}} - (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}}} \le  p^{-c_{\varphi}}, %\norm{\Sigma^{(2)}},
			\end{align}
			where $a_1$ and $a_2$ are the solutions of the following self-consistent equations
			\begin{align}
				a_1 + a_2 = 1- \frac{p}{n_1 + n_2},\quad a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} = \frac{n_1}{n_1 + n_2}. \label{eq_a12extra}
			\end{align}
	\noindent (ii) The bias equation \eqref{eq_bias_2task} satisfies the following limit with high probability for any fixed vector $w\in\real^p$ with unit norm,
			\begin{align}\label{lem_cov_derv_eq}
				\bigabs{w^{\top} \Sigma^{(1)} \bigbrace{(n_1+n_2)^2\hat{\Sigma}^{-1} \Sigma^{(2)} \hat{\Sigma}^{-1} - {\Sigma^{(2)}}^{-1/2} V {\frac{a_3 \Lambda^2 + (a_4 + 1)\id}{(a_1 \Lambda^2 + a_2\id)^2}} V^{\top} {\Sigma^{(2)}}^{-1/2}} \Sigma^{(1)} w} \le  p^{-c_{\varphi}},
			\end{align}
				where $a_{3}$ and $a_4$ are the solutions of the following self-consistent equations % with $b_k = \frac1{p}\sum_{i=1}^p \frac{\lambda_i^{2k}} {(\lambda_i^2 a_1 + a_2)^2}$, for $k = 0, 1, 2$:
			\begin{align}\label{eq_a34extra}
				a_3 + a_4 = \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2}, \ \ 
				a_3 + \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 (a_2 a_3-a_1 a_4 )}{(\lambda_i^2 a_1 + a_2)^2} = \frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^{2}}.
%				\left(\frac{\rho_1}{a_1^{2}} -  b_2  \right)\cdot  a_3 -  b_1 \cdot  a_4 = b_1,\quad \left(\frac{\rho_2}{a_2^{2}}-  b_0\right)\cdot  a_4 - b_1 \cdot  a_3
%				= b_0.
			\end{align}
\end{theorem}
Our result extends Fact \ref{fact_tr} on the inverse of a single sample covariance matrix.
When $n_1$ is zero, we solve equation \eqref{eq_a12extra} and obtain $a_1 = 0$ and $a_2 = (n_2-p) / n_2$, and apply these solutions to equation \eqref{lem_cov_shift_eq}.

@@

\paragraph{Bias asymptotic limit.}
For the bias limit in equation \eqref{lem_cov_derv_eq}, we show that it is governed by the derivative of $(W - z\id)^2$ with respect to $z$ at $z = 0$.
First, we reduce the empirical bias term in equation \eqref{lem_cov_derv_eq} to $W$
\begin{align}\label{calculate G'}
	(n_1 + n_2)^2 \hat{\Sigma}^{-1}\Sigma^{(2)}\hat{\Sigma}^{-1} = {\Sigma^{(2)}}^{-1/2} V W^{-2} V^{\top} {\Sigma^{(2)}}^{-1/2}.
\end{align}
Let $\cal G(z):=(W-z\id )^{-1}$ denote the resolvent of $W$.
Our key observation is that $\frac{\dd{\cal G(z)}}{\dd z} =  \cal G^2(z)$.
Hence, provided that limit of $(W - z\id)^{-1}$ is $(a_1(z) \Lambda^2 + (a_2(z) - z) \id)^{-1}$ near $z = 0$, the limit of $\frac{\dd{\cal G(0)}}{\dd z}$ satisfies that
\begin{align}\label{cal G'0}
	\frac{\dd \cal G(0)}{\dd z} \to \frac{-\frac{\dd a_1(0)}{\dd z}\Lambda^2 - (\frac{\dd a_2(0)}{\dd z} - 1)\id}{(a_{1}(0)\Lambda^2 + a_{2}(0)\id_p)^2}.
\end{align}
To find the derivatives of $a_1(z)$ and $a_2(z)$, we take the derivatives on both sides of equation \eqref{selfomega_a}.
%\begin{align*}
%	\frac{\dd a_1(z)}{\dd z} + \frac{\dd a_2(z)}{\dd z} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{1}{\lambda_i^2 a_1 + a_2},
%	\frac{\dd a_1(z)}{\dd z} + \frac{1}{n_1 + n_2}\sum_{i=1}^p \frac{\lambda_i^2 (a_1'(z) a_2 - a_2'(z) a_1)}{(\lambda_i^2 a_1 + a_2)^2} = -\frac{1}{n_1 + n_2} \sum_{i=1}^p \frac{\lambda_i^2 a_1}{(\lambda_i^2 a_1 + a_2)^2}
%\end{align*}
Let $a_3 = - \frac{\dd a_1(0)}{\dd z}$ and $a_4 = - \frac{\dd a_2(z)}{\dd z}$.
%then taking implicit differentiation of equation \eqref{selfomega_a}
One can verify that $a_3$ and $a_4$ satisfy the self-consistent equations \eqref{eq_a34extra} (details omitted).
Applying equation \eqref{cal G'0} to equation \eqref{calculate G'}, we obtain the asymptotic limit of the bias term.
%\begin{align}
%& n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}) }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
%&\approx  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\frac{a_3\Lambda^2 +(1+ a_4)\id_p}{(a_{1}\Lambda^2 + a_{2}\id)^2}V^\top \Sigma_2^{-1/2}\Sigma_1^{1/2}w= w^{\top} \Pi_\bias w, \label{calculatePibias}
%\end{align}
%where in the last step we used $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $V \Lambda^2 V^\top=M^\top M$. This concludes equation \eqref{lem_cov_derv_eq}.

As a remark, in order for $\frac{\dd \cal G(z)}{\dd z}$ to stay close to its limit at $z = 0$, we not only need to find the limit of $\cal G(0)$, but also the limit of $\cal G(z)$ within a small neighborhood of $0$.
This is why we consider $W$'s resolvent for a general $z$. %as opposed to the Stieljes transform of its empirical spectral distribution.

\paragraph{Schur complement and self-consistent equations.}
We briefly describe how to derive the matrix limit $\Gi(z)$.
First, we consider the special case where $Z^{(1)}$ and $Z^{(2)}$ are both multivariate Gaussian random matrices.
By rotational invariance, we have that $Z^{(1)} U$ and $Z^{(2)} V$ are still multivariate Gaussian random matrices.
Next, we use the Schur complement formula to deal with the resolvent $G(z)$.
We show that $G(z)$'s diagonal entries satisfy a set of self-consistent equations in the limit, leading to equation \eqref{selfomega_a}.
On the other hand, $G(z)$'s off-diagonal entries are approximately zero using standard concentration bounds.
Finally, we extend our result to general random matrices under the finite $\varphi$-th moment condition.
We prove an anisotropic local law using recent developments in random matrix theory \cite{erdos2017dynamical,Anisotropic}.
A complete proof of Theorem \ref{thm_main_RMT} can be found in Appendix \ref{appendix RMT}.

### 2020-10-11 arxiv/experiments.tex
fix missing ref
sec 3

\section{Numerical Results}

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/same_covariates.eps}
	\caption{Validating Example \ref{ex_same_cov} in the same covariates case: our estimated loss (solid line) matches the empirical loss (dotted line) accurately for various task-specific variance $d$ and sample size $n$ settings. The feature dimension $p$ is $200$, and noise variance $\sigma^2$ is $1/4$.}
	\label{fig_same_cov}
\end{figure}

@@

\begin{figure}%[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\vspace{0pt}
		\includegraphics[width=0.6\textwidth]{figures/ratio_norm_3_pairs_lstm.pdf}
		\caption{HPS vs. STL}
		\label{fig_ab_data}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\vspace{0pt}
		\includegraphics[width=0.6\textwidth]{figures/ratio_alignment_norm_diff_all.pdf}
		\caption{HPS vs. covariance alignment}
		\label{fig_ab_cov}
	\end{subfigure}
	\caption{Comparing hard parameter sharing (HPS) to single-task learning (STL) and a covariance alignment approach proposed by \citet{WZR20}:
	In Figure \ref{fig_ab_data}, we observe that for multiple example task pairs, increasing task one's sample size improves task two's prediction accuracy initially, but hurts eventually -- a phenomenon similar  to Figure \ref{fig_size}.
	In Figure \ref{fig_ab_cov}, we observe that as task one's sample size increases, covariance alignment improves more over HPS.}
	\label{fig_text}
%	\begin{minipage}[t]{0.4\textwidth}
%		\centering
%		\vspace{0pt}
%		\begin{tabular}{c c c}
%		\toprule
		% \multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\begin{minipage}{1.2in}\begin{center}
		% Sentiment\\ analysis\end{center}\end{minipage}} \\
%		\multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\bf Sentiment analysis} \\
		% \cmidrule(lr){2-3}
%		& all tasks & w/o TREC \\
%		\midrule
%		{\bf MLP}  & 31\% & 29\% \\
%		{\bf LSTM} & 35\% & 34\% \\
%		{\bf CNN}  & 30\% & 28\% \\
%		\bottomrule
%		\end{tabular}
%		\captionof{table}{Finding the best sample ratio via a progressive training schedule.}
%		\label{tab:taskonomy}
%	\end{minipage}
\end{figure}

>>>>

\section{Numerical Results}\label{sec_exp}

@@

\begin{figure}[!t]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/same_covariates.eps}
	\caption{Validating Example \ref{ex_same_cov} in the same covariates case: our estimated loss (solid line) matches the empirical loss (dotted line) accurately for various task-specific variance $d^2$ and sample size $n$ settings. The feature dimension $p$ is $200$, and noise variance $\sigma^2$ is $1/4$.}
	\label{fig_same_cov}
\end{figure}

@@

\begin{figure}%[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\vspace{0pt}
		\includegraphics[width=0.8\textwidth]{figures/fig3a.pdf}
		\caption{HPS vs. STL}
		\label{fig_ab_data}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\vspace{0pt}
		\includegraphics[width=0.8\textwidth]{figures/fig3b.pdf}
		\caption{HPS vs. covariance alignment}
		\label{fig_ab_cov}
	\end{subfigure}
	\caption{Comparing hard parameter sharing (HPS) to single-task learning (STL) and a covariance alignment approach proposed by \citet{WZR20}:
	In Figure \ref{fig_ab_data}, we observe that for multiple example task pairs, increasing task one's sample size improves task two's prediction accuracy initially, but hurts eventually -- a phenomenon similar  to Figure \ref{fig_size}.
	In Figure \ref{fig_ab_cov}, we observe that as task one's sample size increases, covariance alignment improves more over HPS.}
	\label{fig_text}
%	\begin{minipage}[t]{0.4\textwidth}
%		\centering
%		\vspace{0pt}
%		\begin{tabular}{c c c}
%		\toprule
		% \multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\begin{minipage}{1.2in}\begin{center}
		% Sentiment\\ analysis\end{center}\end{minipage}} \\
%		\multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\bf Sentiment analysis} \\
		% \cmidrule(lr){2-3}
%		& all tasks & w/o TREC \\
%		\midrule
%		{\bf MLP}  & 31\% & 29\% \\
%		{\bf LSTM} & 35\% & 34\% \\
%		{\bf CNN}  & 30\% & 28\% \\
%		\bottomrule
%		\end{tabular}
%		\captionof{table}{Finding the best sample ratio via a progressive training schedule.}
%		\label{tab:taskonomy}
%	\end{minipage}
\end{figure}
